Bayesian Sparsification of $\cplx$-valued networks

### 2019-11-06 paper/paper.tex
compactified the last part of the VD section

For a particular family of posterior approximation, variational inference can be used
as a regularization method \cite{kingma_variational_2015} and as model sparsification
technique \cite{molchanov_variational_2017}. \cite{kingma_variational_2015} propose
Variational Dropout, by studying Binary dropout, DropConnect and Gaussian dropout,
\cite{hinton_improving_2012,wan_regularization_2013}, \cite{srivastava_dropout_2014},
and \cite{wang_fast_2013} through the lens of variational inference methods. The study
argues that the multiplicative noise, introduced by dropout, induces a distribution
equivalent to a fully factorized variational posterior of the form $
  q_\theta(\omega) = \prod_j q_{\theta}(\omega_j)
$, where $q_{\theta}(\omega_j)$ is $\omega_j = \mu_j \xi_j$ with $\xi_j \sim p(\xi_j)$
iid from some $p_\xi$. Binary dropout with rate $p \in (0, 1)$ uses $
  p_\xi
    = \mathcal{B}\bigl(
      \{0, \tfrac1{1-p}\}, 1-p
    \bigr)
$, whereas in Gaussian dropout $
  \xi \sim \mathcal{N}(1, \alpha)
$ for $\alpha = \tfrac{p}{1-p}$. This suggests making $\alpha$ a variational parameter
and optimizing it in \eqref{eq:elbo_general} with an appropriate penalty term. Flexibility
in dependence of $q_\theta$ on $\alpha$ allows structured dropout with shared $\alpha$
across groups of parameters or sparsifying dropout, where the rate $\alpha$ is optimized
for each individual parameter, \cite{molchanov_variational_2017}.

Variational Dropout with mean field Gaussian variational approximation is further developed
in \cite{molchanov_variational_2017} and \cite{kharitonov_variational_2018}. The first paper
studies variational dropout from a model sparsification perspective and improves on the
results of \cite{kingma_variational_2015} in two directions: the authors propose a more
accurate approximation to the KL-divergence term in \eqref{eq:elbo}, and introduce additive
noise parameterization for the factors of $q_\theta(\omega)$, $\omega = W$.

Under the original $(\mu, \alpha)$ parameterization the stochastic output of a linear
layer $
  y = W^\top x + b
$ with $
  W \in \mathbb{R}^{n\times m}
$ and independent $
  W_{ij} \sim \mathcal{N}(\mu_{ij}, \alpha_{ij} \mu_{ij}^2)
$ is equivalent in distribution to $
  y \sim \prod_i \mathcal{N}\Bigl(
      \sum_j \mu_{ij} x_j + b_i,
      \sum_j \alpha_{ij} \mu_{ij}^2 x_j^2
  \Bigr)
$. \cite{molchanov_variational_2017} argue that for the purposes of model sparsification
this parameterization should be abandoned in favour of an equivalent $(\mu, \sigma^2)$ with
the score $\alpha$ calculated using $\tfrac{\sigma^2}{\mu^2}$ when needed. This makes $
  \nabla_\mu \tilde{\mathcal{L}}
$, which is particularly important for sparsification, independent of the noise injected
by the trick.

\cite{kingma_variational_2015} and \cite{molchanov_variational_2017} consider a factorized
prior $\pi(\omega)$ with improper log-uniform factors $
  \pi(\omega_j) \propto \lvert \omega_j \rvert^{-1}
$. This unravels the divergence term in \eqref{eq:elbo} into a sum of the following terms:
\begin{equation} \label{eq:improper-kl-div-real}
  KL(q_\theta(\omega) \| \pi(\omega))
    \propto
      \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
        \log{\Bigl\lvert \tfrac1{\sqrt{\alpha}} + \varepsilon \Bigr\rvert^2}
      % \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
      %   \log{\bigl\lvert 1 + \sqrt{\alpha} \varepsilon \bigr\rvert^2}
      % - \frac12 \log{\alpha}
  \,.
\end{equation}
The expectation in \eqref{eq:expect-improper-term-cplx} is half the $\log$-moment of a non-central
$\chi^2$ distribution with shape $1$ and non-centrality $\lambda = \tfrac1{\alpha}$. In general,
there is no analytically tractable expression for log-moment of a $\chi^2_m$ with odd degrees
of freedom. To circumvent this, \cite{kingma_variational_2015} approximate the expectation by
a non-linear polynomial regression for truncated $\alpha \in (0, 1)$. This approximation was
substantially refined in \cite{molchanov_variational_2017}, both in term of the domain of scores
$\alpha$ and accuracy. The paper considered regression with sigmoid and soft-plus basis functions
fit to the Monte-Carlo estimator of \eqref{eq:improper-kl-div-real} for $\alpha$ varying over
a fine logarithmic grid. In sec.~\ref{sub:real-chisq-grad} we verify the approximation and
compare the approximate derivative to the exact derivative of \eqref{eq:improper-kl-div-real}
with respect to $\log \alpha$.
% Since gradient based methods rely on unbiased gradient estimates

In another variant of Bayesian dropout, called Automatic Relevance Determination, is developed
in the \cite{kharitonov_variational_2018}. The key difference from Variational Dropout is the
choice of prior: the log-uniform factor $\pi(\omega_{ij})$ is replaced by a proper Gaussian
prior $\mathcal{N}(0, \tau^{-1}_{ij})$ with learnable precision $\tau_{ij} > 0$. This is knowns
as the Empirical Bayes method, \cite{citation_needed}, which fits a prior distribution while
performing variational inference. The optimal precision parameter is found in the closed form
by minimizing the KL-divergence term holding variational parameters $\theta$ fixed.

Second contribution of \cite{kingma_variational_2015} is the analysis of the \textit{local
reparameterization trick}, proposed in \cite{wang_fast_2013} to speed up Gaussian dropout,
as a gradient variance reduction technique for \eqref{eq:elbo}. This trick utilizes the
closure of Gaussians under affine transformations, thereby translating uncertainty from the
global parameter noise to local noise in intermediate outputs within the network.
%
\cite{kingma_variational_2015} show, that for the case of fully factorized Gaussian
posterior approximations, the trick decorrelates ELBO estimates within the minibatch,
making the gradient estimator $\nabla_\theta \tilde{\mathcal{L}}$ more statistically
efficient.

>>>>

For a particular family of posterior approximation, variational inference can be used
as a regularization method \cite{kingma_variational_2015} and as model sparsification
technique \cite{molchanov_variational_2017}. \cite{kingma_variational_2015} propose
Variational Dropout, by studying Binary dropout, DropConnect and Gaussian dropout,
\cite{hinton_improving_2012,wan_regularization_2013}, \cite{srivastava_dropout_2014},
and \cite{wang_fast_2013} through the lens of variational inference methods. The study
argues that the multiplicative noise, introduced by dropout, induces a distribution
equivalent to a fully factorized variational posterior of the form $
  q_\theta(\omega) = \prod_j q_{\theta}(\omega_j)
$, where $q_{\theta}(\omega_j)$ is $\omega_j = \mu_j \xi_j$ with $\xi_j \sim p(\xi_j)$
iid from some $p_\xi$. Binary dropout with rate $p \in (0, 1)$ uses $
  p_\xi
    = \mathcal{B}\bigl(
      \{0, \tfrac1{1-p}\}, 1-p
    \bigr)
$, whereas in Gaussian dropout $
  \xi \sim \mathcal{N}(1, \alpha)
$ for $\alpha = \tfrac{p}{1-p}$. This suggests making $\alpha$ a variational parameter
and optimizing it in \eqref{eq:elbo_general} with an appropriate penalty term. Thus,
\cite{kingma_variational_2015} propose \textit{Variational Dropout} with mean field Gaussian
approximation $
  q_\theta(\omega)
    = \prod_j \mathcal{N}(\mu_j, \alpha_j \mu_j^2)
$ and factorized log-uniform prior $\pi(\omega)$ with $
  \pi(\omega_j) \propto \lvert \omega_j \rvert^{-1}
$. This unravels the divergence term in \eqref{eq:elbo} into a sum of the following terms:
\begin{equation} \label{eq:improper-kl-div-real}
  KL(q_\theta(\omega) \| \pi(\omega))
    \propto
      \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
        \log{\Bigl\lvert \tfrac1{\sqrt{\alpha}} + \varepsilon \Bigr\rvert^2}
      % \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
      %   \log{\bigl\lvert 1 + \sqrt{\alpha} \varepsilon \bigr\rvert^2}
      % - \frac12 \log{\alpha}
  \,.
\end{equation}
In general, there is tractable expression for log-moment of a non-central $\chi^2_m$ random
variable with odd $m$, that sits under the expectation in \eqref{eq:expect-improper-term-cplx}.
To circumvent this, \cite{kingma_variational_2015} approximate the expectation by a non-linear
polynomial regression for truncated $\alpha \in (0, 1)$. This approximation was refined in
\cite{molchanov_variational_2017}, both in terms of the domain of $\alpha$ and accuracy. The
paper considered regression with sigmoid and soft-plus basis functions fit to the Monte-Carlo
estimator of \eqref{eq:improper-kl-div-real} for $\alpha$ varying over a fine logarithmic grid.
In sec.~\ref{sub:real-chisq-grad} we verify the approximation and confirm its accuracy, by
comparing its derivative to the exact derivative of \eqref{eq:improper-kl-div-real} with
respect to $\log \alpha$.
% (!) good for SGVB: gradient-based methods require unbiased gradient estimates

Flexibility in dependence of $q_\theta$ on $\alpha$ allows structured dropout with shared
$\alpha$ across groups of parameters $\omega$. \cite{molchanov_variational_2017} study
Variational Dropout from a model sparsification perspective, optimizing $\alpha$ for each
individual parameter. Another variant of Bayesian dropout, called Automatic Relevance
Determination, is developed in \cite{kharitonov_variational_2018}. The key difference from
Variational Dropout is the choice of prior: the log-uniform factor $\pi(\omega_{ij})$ is
replaced by a proper Gaussian prior $
  \mathcal{N}(0, \tau^{-1}_{ij})
$ with learnable precision $\tau_{ij} > 0$. This is known as the Empirical Bayes method,
\cite{citation_needed}, which fits a prior distribution while performing variational inference.
The optimal precision parameter is found in the closed form by minimizing the KL-divergence
term holding variational parameters $\theta$ fixed.

Another contribution of \cite{kingma_variational_2015} is the analysis of the \textit{local
reparameterization trick}, proposed in \cite{wang_fast_2013} to speed up Gaussian dropout,
as a gradient variance reduction technique for \eqref{eq:elbo}. This trick utilizes the
closure of Gaussians under affine transformations, thereby translating uncertainty from the
global parameter noise to local noise in intermediate outputs within the network.
%
The stochastic output of a linear layer $
  y = W^\top x + b
$ with $
  W \in \mathbb{R}^{n\times m}
$ and independent $
  W_{ij} \sim \mathcal{N}(\mu_{ij}, \alpha_{ij} \mu_{ij}^2)
$ is equivalent in distribution to
\begin{equation} \label{eq:r-gauss-trick}
    y \sim \prod_i \mathcal{N}\Bigl(
          \sum_j \mu_{ij} x_j + b_i,
          \sum_j \alpha_{ij} \mu_{ij}^2 x_j^2
      \Bigr)
    \,.
\end{equation}
%
\cite{kingma_variational_2015} show, that for the case of fully factorized Gaussian
posterior approximations, the trick decorrelates ELBO estimates within the minibatch,
making the gradient estimator $\nabla_\theta \tilde{\mathcal{L}}$ more statistically
efficient. \cite{molchanov_variational_2017} propose an improvement by introducing
\textit{additive noise parameterization}, which is relevant to model sparsification.
The paper argues that $(\mu, \alpha)$ parameterization should be abandoned in favour
of an equivalent $(\mu, \sigma^2)$ with the score $\alpha$ calculated using $
  \tfrac{\sigma^2}{\mu^2}
$ when needed. This makes $
  \nabla_\mu \tilde{\mathcal{L}}
$, which is shown in \cite{molchanov_variational_2017} to be independent of the noise,
injected by the trick \eqref{eq:r-gauss-trick}, and further stabilizes gradients in SGVB
\eqref{eq:elbo}.
